---
title: "London Cycle Hires and Weather"
output: html_document
---
London Cycle Hires and Weather
==============================

My goal is to investigate usage of the London cycle hire scheme, and in particular how it varies with the weather. I'm running an analysis from July 2013 to June 2014.

I'm using two data sets here. Daily weather data comes from [Weather Underground](http://www.wunderground.com/history/airport/EGLL/2013/7/1/CustomHistory.html?dayend=30&monthend=6&yearend=2014&req_city=&req_state=&req_statename=&reqdb.zip=&reqdb.magic=&reqdb.wmo=), using the weather station at London Heathrow airport.

(London City Airport is closer to the bike stations that I use, but the data
from that airport reports 0 precipitation on every single day. The data from
Heathrow seems to be more complete, and I expect it to be almost as relevant.)

I collected the cycle hire data myself, over the course of the year, by downloading CSV files from an [unofficial API](http://borisapi.heroku.com) which now appears to be defunct. It has a granularity of about ten minutes. That's about 50,000 entries per docking station for the year, so for this analysis, I'm only using the data from four docking stations near my office.

All data and source code used for this project can be found in the [git repository](https://github.com/ChickenProp/cycle-hire-weather).

```{r load.libs, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
library(ggplot2)
library(reshape2)
library(dplyr)
library(lubridate)
library(GGally)
library(scales)
```

```{r parse.data, echo=FALSE, cache=TRUE}
# First load the weather.

weather <- read.csv('weather.csv')
names(weather) <- c('date', 't.max', 't', 't.min', 'dew.max', 'dew', 'dew.min',
                    'hum.max', 'hum', 'hum.min', 'pres.max', 'pres', 
                    'pres.min', 'vis.max', 'vis', 'vis.min', 'wind.max',
                    'wind', 'gust', 'precip.mm', 'cloud', 'events', 'wind.dir')

# Remove columns that I'm not going to be looking at
weather <- subset(weather, select=c('date', 't.max', 't', 't.min', 'wind.max',
                                    'wind', 'gust', 'precip.mm', 'events'))

weather$date <- as.Date(weather$date)

# The events are Fog, Rain, Snow, and Thunderstorm. Only rain and fog have 
# enough days to be worth looking at (even fog is marginal, at 36).
weather$rain <- grepl('Rain', weather$events)
weather$fog <- grepl('Fog', weather$events)

# I'm going to need times with no dates attached. The way to do that seems to
# be to have datetimes with a constant date.
#   This can parse numbers (seconds into the day), or strings in the form 
# 'HHMM', or POSIXcts.
as.time <- function (ct) {
  if (is.character(ct))
    return(as.time(as.POSIXct(ct, format='%H%M', tz='GMT')))
  return(as.POSIXct(as.numeric(ct) %% 86400, origin = "1970-01-01", tz='GMT'))
}

# It's easier to wrangle the bike files separately and then put them together
# than to put them together before wrangling, because of the prev.* and d.*
# columns.

read.bikes <- function (fname) {
  bikes <- read.csv(fname)

  # Remove columns we won't be looking at. id/lat/long are constant within each
  # file. The others are constant over the whole dataset.
  bikes$id <- NULL
  bikes$lat <- NULL
  bikes$long <- NULL
  bikes$installed <- NULL
  bikes$locked <- NULL
  bikes$temporary <- NULL

  # Rename and restructure
  names(bikes) <- c('updated', 'name', 'num.bikes', 'num.spaces')
  bikes$updated <- as.POSIXct(strptime(bikes$updated, "%Y-%m-%dT%H:%M:%S%z", 
                                       tz='GMT'))

  # Add columns for prev and diff between updates. I feel like there should be
  # a better way to do this, but I haven't found it. (Time series don't seem to
  # work very well.) Only add ones I actually use.
  bikes$prev.updated <- c(as.POSIXct(NA), head(bikes$updated, -1))
  attributes(bikes$prev.updated)$tzone <- 'GMT'
  bikes$prev.num.bikes <- c(NA, head(bikes$num.bikes, -1))

  bikes$d.updated <- with(bikes, as.numeric(updated - prev.updated))
  bikes$d.num.bikes <- with(bikes, num.bikes - prev.num.bikes)
  
  # The timestamps are given in absolute time, but we're interested in clock
  # time, which was +1hr before 2013-10-27 and from 2014-03-30. Note that we
  # need to do this *after* creating d.updated, or d.updated might be wrong
  # around timezone boundaries.
  no.bst <- strptime(c('2013-10-27 01:00:00', '2014-03-30 02:00:00'),
                     '%Y-%m-%d %H:%M:%S', tz='GMT')
  adj.bst <- function (ct)
    ct + ifelse(no.bst[1] <= ct & ct <= no.bst[2], 0, 3600)
  bikes$updated <- adj.bst(bikes$updated)
  bikes$prev.updated <- adj.bst(bikes$prev.updated)

  # Add columns derived from existing ones.
  bikes$updated.time <- as.time(bikes$updated)
  bikes$updated.date <- as.Date(bikes$updated)
  bikes$num.docks <- with(bikes, num.bikes+num.spaces)

  bikes$prev.updated.time <- as.time(bikes$prev.updated)
  bikes$prev.updated.date <- as.Date(bikes$prev.updated)
  bikes$prev.num.docks <- c(NA, head(bikes$num.docks, -1))

  bikes$weekday <- wday(bikes$updated.date, label=T)
  bikes$is.weekday <- wday(bikes$updated.date) %in% seq(2,6)

  # Remove the first row, to avoid NAs in prev.* and d.* columns.
  return(bikes[-1,])
}

bikes <- rbind(read.bikes('bikes-sp.csv'),
               read.bikes('bikes-hh.csv'),
               read.bikes('bikes-bp.csv'),
               read.bikes('bikes-es.csv'))

# The subset is justified later on.
bikes.all <- merge(x=bikes, y=weather, by.x='updated.date', by.y='date')
bikes <- subset(bikes.all, d.updated <= 15)
```

```{r at.time, echo=FALSE, cache=TRUE}
# I also want to look at the number of bikes available at specific times. Since 
# I only have snapshots, I'm going to take the first observation after that 
# time on any given day. Here's a function that lets me do that. It takes a 
# time in formats accepted by as.time, and returns a boolean vector to select 
# the appropriate entries from bikes.
at.time <- function(time) {
  time <- as.time(time)
  with(bikes, ifelse(prev.updated.date == updated.date, 
                     prev.updated.time < time & time <= updated.time,
                     prev.updated.time < time | time <= updated.time))
}
```

Exploring the weather data
--------------------------

#### Temperature

```{r temp.1v, echo=FALSE, cache=TRUE}
temps <- melt(weather, id.vars='date', measure.vars=c('t.min','t','t.max'))
ggplot(temps, aes(x=variable, y=value)) + geom_boxplot()
ggplot(temps, aes(x=value, fill=variable)) + geom_density(alpha=0.3)
```

These variables measure the minimum, average, and maximum daily temperatures.
The graphs all look similar, and overlap a lot. The shape is a little
surprising, as I didn't expect the density graphs to be bimodal. It could
potentially be caused by significant differences between summer and winter, with
an abrupt shift between the two.

#### Rainfall

```{r rain.1v, echo=FALSE, cache=TRUE}
ggplot(weather, aes(x=rain)) + geom_histogram()
ggplot(weather, aes(x=precip.mm)) + geom_histogram(binwidth=0.2)
```

According to the `rain` column, There are over 225 rainy days and only about 125 non-rainy days. But by far the most common bin for `precip.mm` is the leftmost one. Table of values of `precip.mm`:

```{r rain.1v.table, echo=FALSE, cache=TRUE}
table(weather$precip.mm)
```

Although more than half of observations have `rain == TRUE`, more than half of them also have `precip.mm == 0`, which needs more investigation. Rainfall as measured by `precip.mm` versus as measured by `rain`:

```{r rain.precip.2v, echo=FALSE, cache=TRUE}
ggplot(weather, aes(y=precip.mm, x=rain)) + geom_boxplot()
ggplot(weather, aes(x=precip.mm, fill=rain, color=rain)) + 
  geom_histogram(position='dodge', binwidth=0.1)
```

The two measures don't always agree. Sometimes `rain` is false but `precip.mm` is nonzero; and often `rain` is true but `precip.mm` is zero. Neither of those is surprising individually: if `rain` is only counted when the rainfall exceeds a certain threshold, then that threshold could be large (giving false/nonzero) or small (giving true/zero). But the combination suggests that that isn't what's going on, and I don't know what is.

This table counts the anomalies by turning `precip.mm` into a boolean zero/nonzero (false/true) and comparing it to `rain`:

```{r rain.precip.2v.table, cache=TRUE}
table(weather$rain, weather$precip.mm != 0)
```

There are 88 instances of true/zero, 9 instances of false/nonzero, but the cases where they agree are the most common.

I find `precip.mm` to me more plausible here. I feel like fewer than half of days are rainy. [This website](http://www.london.climatemps.com/) agrees with me, saying that on average, 164 days out of the year are rainy (`rain` - 237, `precip.mm` - 158).

#### Wind

```{r wind.1v, echo=FALSE, warning=FALSE, cache=TRUE}
winds <- melt(weather, id.vars='date',
              measure.vars=c('wind','wind.max','gust'))
ggplot(winds, aes(x=variable, y=value)) + geom_boxplot()
ggplot(winds, aes(x=value, fill=variable)) + geom_density(alpha=0.3)
```

These three measures of wind speed are all averages. `wind` is simply the average wind speed over a day. `wind.max` is the daily maximum of the average wind speed over a short time period (I think one minute). `gust` is the same thing, but with a shorter time period (I think 14 seconds).

Unlike with temperature, the three measures look different. All are right-skewed, although `gust` looks less so. There are several outliers (the isolated points on the box plots), and the quartiles don't overlap. The minimum gust speed (about 24) is almost as high as the median `wind.max`.

Exploring the bike data
-----------------------

#### Time between updates

```{r dt.1v, cache=TRUE}
# Histogram of d.updated, on a log-y scale. The ..count..+1 avoids ugly blocks
# hanging down where there are no entries in a bin.
ggplot(bikes.all, aes(x=d.updated, y=..count..+1)) + 
  geom_histogram(binwidth=20) +
  scale_y_log10()
```

There are a few outliers here. Not all the lines are visible due to rendering artifacts, but above 5000, we only have five entries:

```{r dt.1v.right, cache=TRUE}
bikes.all[bikes.all$d.updated >= 5000, c('name', 'prev.updated', 'updated')]
```

The first four of these happened when my collection script broke and I failed to realize it. The other occurred when Southampton Place was taken out of service temporarily.

Let's zoom in on the lower ones:

```{r dt.1v.left, cache=TRUE}
ggplot(bikes.all[bikes.all$d.updated < 5000,], 
       aes(x=d.updated, y=..count..+1)) +
  geom_histogram(binwidth=5) +
  scale_y_log10()

# The times of entries with between 2000 and 5000 minutes between updates.
bikes.all[bikes.all$d.updated >= 2000 & bikes.all$d.updated < 5000,
          c('name', 'prev.updated', 'updated')]

# Now entries with less than 60 minutes between updates, no longer on a 
# log scale.
ggplot(bikes.all[bikes.all$d.updated < 60,], aes(x=d.updated)) + 
  geom_histogram(binwidth=1)
```

Then there are a several instances where the time between updates is unusually large, on the order of hours or days. It looks like these happened to all stations simultaneously, suggesting problems with either my collection script or the API, rather than problems with individual locations.

But in the vast majority of cases, updates are approximately ten minutes apart. This encourages me to take a subset of the data (`bikes.all` -> `bikes`), considering only entries with `d.updated` less than 15 minutes. This eliminates many outliers in future graphs.

#### Date and time of update

```{r date.time.1v, echo=FALSE, cache=TRUE}
ggplot(bikes.all, aes(x=updated.date)) + 
  geom_histogram(binwidth=1) + 
  scale_x_date()

ggplot(bikes.all, aes(x=updated.time)) + 
  geom_histogram(binwidth=600) +
  scale_x_datetime(labels=date_format('%H:%M'))
```

All times of day are approximately equally represented to within ten minutes, which is good. There are five noticeable troughs preceeded by spikes, but they probably don't signify much. Dates are a lot less uniform, however. Even apart from the ten-day period where my script was broken, many days have significantly fewer updates than typical, and some have none at all.

#### Number of days spent with a given number of active docks

```{r ndocks.time.2v, echo=FALSE, warning=FALSE, cache=TRUE}
ggplot(bikes, aes(x=prev.num.docks, weight=d.updated/60/24)) +
  geom_bar(binwidth=1) + scale_x_discrete() +
  facet_wrap(~name)
```

It was common for every station to report less than a full complement of docks. At least two had a full complement for less than half the time (High Holborn and Bury place are unclear in that respect). This isn't surprising, since a bike reported as defective will be locked in, using up a slot but not being available for hire.

#### Journeys taken throughout the year

```{r date.journeys.2v, cache=TRUE}
ggplot(bikes, aes(x=updated, y=abs(d.num.bikes))) + 
  geom_jitter(alpha=0.05) + stat_smooth()
```

The time of year makes very little difference to the number of rides. There appears to be a slight sinusoidal relationship, but it's very weak. (I didn't do a PMCC test because that assumes that any relationship is linear, which we would naively expect not to be the case here, and also doesn't look true from the graph.)

#### Journeys by weekday

```{r weekday.journeys.2v, warning=FALSE, cache=TRUE, echo=FALSE}
ggplot(bikes, aes(x=weekday, weight=abs(d.num.bikes))) + 
  geom_bar() +
  ggtitle('Total absolute change in num.bikes')

ggplot(bikes, aes(x=weekday, y=num.bikes)) + 
  geom_boxplot() +
  ggtitle('Number of bikes available')

ggplot(bikes, aes(x=num.bikes/num.docks)) + 
  geom_density() + 
  facet_wrap(~weekday) +
  ggtitle('Station fullness')
```

Fewer journeys are taken on weekends. The median number of bikes available doesn't change much throughout the week (5 on monday and friday, 4 on other days), but the distribution does. Saturday and Sunday have noticeably different shapes to the others. They have a single peak, while weekdays are somewhat bimodal, with a small peak where the station is full (probably when people are arriving at work).

(Since the stations have different numbers of docks, I did a graph of fullness rather than of number of bikes. The density plot doesn't show peaks exactly at 0 and 1 because of how the density window works, but histograms of num.bikes and num.spaces show that that's where they are. It would be difficult to use a histogram for this graph because there's no sensible binwidth.)

#### Change in number of bikes between updates

```{r bikes.prevbikes.name.mv, warning=FALSE, cache=TRUE, echo=FALSE}
ggplot(bikes, aes(x=num.bikes, y=prev.num.bikes, color=name)) + 
  geom_jitter(alpha=0.05)

cor.test(bikes$num.bikes, bikes$prev.num.bikes)
```

There's very strong correlation between the number of bikes in adjacent entries. This is as expected, especially given what we saw about `d.num.bikes` previously. The colors here don't show any particular station-dependent trends.

#### Number of bikes at any given time

```{r bikes.time.name.mv, warning=FALSE, cache=TRUE, echo=FALSE}
ggplot(bikes, aes(x=updated.time, y=num.bikes, color=name)) + 
  geom_jitter(alpha=0.05, shape=1) +
  stat_smooth()
```

The correlation also looks strong between the number of bikes at each station at any given time. Since they're all close to each other, that's not surprising. The time is a big factor, with large numbers of bikes in the stations during office hours, and few numbers in the evening and early morning. There's a slight dip around 1pm, which could be related to people using them on their lunch breaks.

This graph gives an overview of global trends, but I mostly use the bikes at specific times. We can zoom in on those:

#### Number of slots available at 0930

(when I'm trying to arrive at work)

This is a proportional frequency plot: within each facet of the graph, the heights of the bins add up to 1. Only weekdays are considered.

```{r slots.0930.2v, echo=FALSE, cache=TRUE}
# Every entry in bikes gets a new column, which is the reciprocal number of 
# entries with the same value for name. Then we stack bars of that height on
# top of each other.
ggplot(bikes[at.time('0930') & bikes$is.weekday,] %>% 
         group_by(name) %>% 
         mutate(frequency=1/length(name)),
       aes(x=num.spaces, y=frequency)) +
  geom_bar(stat='identity', binwidth=1) +
  facet_wrap(~name)
```

About 40% of the time, Earnshaw street has no spaces. That's actually less than I'd realized. It's directly outside my office, and I haven't even been checking it because I'd assumed it was always full.

#### And at 0940

(in case I'm running late)

```{r slots.0940.2v, echo=FALSE, cache=TRUE}
ggplot(bikes[at.time('0940') & bikes$is.weekday,] %>% 
         group_by(name) %>% 
         mutate(frequency=1/length(name)),
       aes(x=num.spaces, y=frequency)) +
  geom_bar(stat='identity', binwidth=1) +
  facet_wrap(~name)
```

If I'm late, I have slightly less chance of finding a docking station, but not much less.

Combining the two
-----------------

#### Journeys taken on rainy vs. non-rainy days

Here, `rain` is the original variable in the dataset, and `rain2` simply measures whether `precip.mm` is nonzero. We have graphs looking at `d.num.bikes` on each type of day, and tables comparing its mean absolute value.

```{r rain.rain2.journeys.2v, warning=FALSE, cache=TRUE, echo=FALSE}
ggplot(group_by(bikes, rain) %>% mutate(recip.count=1/length(rain)),
       aes(x=d.num.bikes, weight=recip.count)) + 
  geom_bar() +
  scale_x_discrete() +
  facet_wrap(~rain) +
  ggtitle('rain')

# Mean absolute change in number of bikes, depending on value of rain.
group_by(bikes, rain) %>% summarise(mean(abs(d.num.bikes)))

# Same thing, but factored over whether precip.mm is nonzero.
bikes$rain2 <- bikes$precip.mm != 0
ggplot(group_by(bikes, rain2) %>% mutate(recip.count=1/length(rain2)),
       aes(x=d.num.bikes, weight=recip.count)) + 
  geom_bar() +
  scale_x_discrete() +
  facet_wrap(~rain2) +
  ggtitle('rain2')

group_by(bikes, rain2) %>% summarise(mean(abs(d.num.bikes)))

# Mean absolute change, depending on values of both rain and rain2.
group_by(bikes, rain, rain2) %>% summarise(mean(abs(d.num.bikes)))
```

Earlier I said I feel like `precip.mm` is more accurate than `rain`. Despite that, `rain` seems to be capturing something that `precip.mm` doesn't, because bike usage responds slightly more to it. This would seem to suggest that days where `rain` is true but `precip.mm` is zero have less bike usage than average; and indeed this is what we see.

Taking `rain` to be our measure, slightly over 70% of observations had no bikes added or removed on rainy days, and slightly under 70% on non-rainy days. The mean absolute difference is about 25% higher on non-rainy days.

#### Foggy versus non-foggy days

```{r fog.journeys.2v, warning=FALSE, cache=TRUE, echo=FALSE}
ggplot(group_by(bikes, fog) %>% mutate(recip.count=1/length(fog)),
       aes(x=d.num.bikes, y=recip.count)) +
  geom_bar(stat='identity') +
  scale_x_discrete() +
  facet_wrap(~fog)

group_by(bikes, fog) %>% summarise(mean(abs(d.num.bikes)))
```

#### Journeys by temperature and wind:

```{r temp.wind.journeys.2v, cache=TRUE, echo=FALSE}
ggplot(bikes, aes(x=t, y=abs(d.num.bikes))) + 
  geom_jitter(alpha=0.05) + stat_smooth()
cor.test(bikes$t, abs(bikes$d.num.bikes))

ggplot(bikes, aes(x=wind, y=abs(d.num.bikes))) + 
  geom_jitter(alpha=0.01) + stat_smooth()
cor.test(bikes$wind, abs(bikes$d.num.bikes))
```

Unlike rain, it seems that fog, wind and temperature make approximately no difference. The mean absolute difference in number of bikes is about the same regardless of fog, and the correlation between that and temperature/wind is close to zero.

#### Number of bikes at any given time, depending on rain:

```{r bikes.time.rain.mv, warning=FALSE, cache=TRUE, echo=FALSE}
ggplot(bikes, aes(x=updated.time, y=num.bikes, color=rain)) +
  geom_jitter(alpha=0.05, shape=1) +
  stat_smooth()
```

Rain reduces the variance, with fewer bikes during office hours and more outside of them.

Reformatting
------------

With the data in the current format, not all the questions we want to ask are easy. For example: how does the number of bikes at one station correlate with another at any given time? I previously said it "looks strong", but that's pretty vague.

To answer questions like that, we need to be somewhat forgiving with our definition of 'any given time'. Updates don't necessarily happen simultaneously, so we need to bin them together.

I'm going to create bins ten minutes wide, and assign every observation to a bin. Then in each bin, we can ask how many bikes were at each station. Using this, we can check correlation between each station:

```{r bin.time, cache=TRUE, echo=FALSE}
# Round updated.time to the nearest ten minutes
tmp <- as.numeric(bikes$updated.time) + 300
bikes$updated.time.bin <- as.time(tmp - tmp%%600)
rm(tmp)

# This gives a slightly different result: the latest :?5:00 time between
# updated.time and prev.updated.time, or NA if there isn't one. (Day boundary
# caveat: If the updates were e.g. at 23:50:00 and  00:10:00, then the bin will
# be 23:55 rather than 00:05.)
#   I'm not sure which version is more appropriate, but there's not much in it,
# so I'm using the other one.

# for (time in seq(300, 60*60*24, 600)) {
#   bikes$updated.time.bin <- 
#     ifelse(at.time(as.time(time)), time, bikes$updated.time.bin)
# }
# bikes$updated.time.bin <- as.time(bikes$updated.time.bin)

# Create wide-format data. The station ordering comes from the factor levels of
# bikes$name, which comes from the order I originally loaded them.
bikes.wide <- dcast(bikes, updated.date + updated.time.bin ~ name,
                    value.var='num.bikes', fun.aggregate=mean)
names(bikes.wide) <- c('date', 'time', 'sp', 'hh', 'bp', 'es')
bikes.wide$total <- with(bikes.wide, sp + hh + bp + es)
```

```{r ggpairs, echo=FALSE, warning=FALSE, cache=TRUE}
ggpairs(select(bikes.wide, sp:es),
        lower=list(params=c(alpha=0.05, position='jitter')))
```

Correlations range between 0.703 and 0.758, and the scatter plots and density histograms all look pretty similar. Does the correlation depend on time? Let's go for 0930, 1800, midnight, and noon.

```{r ggpairs.times, warning=FALSE, cache=TRUE, echo=FALSE}
ggpairs(select(filter(bikes.wide, time == as.time('0930')), sp:es),
        lower=list(params=c(alpha=0.05, position='jitter')), title='0930')

ggpairs(select(filter(bikes.wide, time == as.time('1800')), sp:es),
        lower=list(params=c(alpha=0.05, position='jitter')), title='1800')

ggpairs(select(filter(bikes.wide, time == as.time('0000')), sp:es),
        lower=list(params=c(alpha=0.05, position='jitter')), title='midnight')

ggpairs(select(filter(bikes.wide, time == as.time('1200')), sp:es),
        lower=list(params=c(alpha=0.05, position='jitter')), title='noon')
```

The correlations are almost all lower. That surprised me, but I think it's an example of [Simpson's paradox](http://en.wikipedia.org/wiki/Simpson%27s_paradox). I note that the darkest points in the graph are at midnight, with no bikes in any station much of the time. Bikes are periodically moved in vans to account for anticipated demand; I assume that these stations are emptied most nights to prepare for people coming to work in the morning.

An interesting point is that the weakest correlation on any of the graphs is 0.149, between Earnshaw Street and Bury Place at 1800. But the strongest correlation at a specific time is 0.757, also between those two stations, at 0930.

We also see the density charts sometimes having very different shapes, especially at 0930 and 1800. But this seems to be at least partly to do with the way that `ggpairs` chooses the axes on its density plots. For example, here's 0930:

```{r bikes.0930.density, echo=FALSE, cache=TRUE}
ggplot(bikes[at.time('0930'),], aes(x=num.bikes)) + 
  geom_density() + 
  facet_wrap(~name, scales='free')
```

The troughs look a lot less significant now.

We can view a histogram of the total number of bikes available at different times:

```{r bikes.time.hists, warning=FALSE, cache=TRUE, echo=FALSE, fig.width=11, fig.height=11}
ggplot(bikes.wide, aes(x=total)) + 
  geom_histogram(binwidth=1) + 
  facet_wrap(~time)
```

We see heavy leftward skews overnight, with much flatter (but somewhat right-skewed) distributions during office hours, and gradual transitions between the two.

We can also check correlation between times more distant than a single tick. If I check the slots available when I leave the house, can I learn how many will be there when I arrive?

```{r cor.0900.0930, warning=FALSE, cache=TRUE, echo=FALSE}
# Take the number of spaces at 0900 and 0930 in each station on each day, in 
# wide format (i.e. both in the same row). If there are multiple entries in a 
# single bin, take the mean number of spaces.
spaces.0900.0930 <- 
  dcast(bikes[bikes$updated.time.bin %in% as.time(c('0900', '0930')),], 
        updated.date + name + rain + is.weekday ~ 
          strftime(updated.time.bin, 'at.%H%M', tz='GMT'),
        value.var='num.spaces', fun.aggregate=mean)

ggplot(spaces.0900.0930, aes(x=at.0900, y=at.0930, color=name)) +
  geom_jitter(alpha=0.2) +
  stat_smooth()

with(spaces.0900.0930, cor.test(at.0900, at.0930))
```

This is good correlation! Does it depend on the rain?

```{r cor.0900.0930.rain, warning=FALSE, cache=TRUE, echo=FALSE}
ggplot(spaces.0900.0930, aes(x=at.0900, y=at.0930, color=rain)) +
  geom_jitter() +
  stat_smooth()

with(spaces.0900.0930, cor.test(at.0900[rain], at.0930[rain]))
with(spaces.0900.0930, cor.test(at.0900[!rain], at.0930[!rain]))
```

Not much, if at all.

We can construct a model

```{r lm, cache=TRUE, echo=FALSE}
summary(lm(at.0930 ~ at.0900, spaces.0900.0930))
```

with an R^2 of 0.79, which is okay. But this isn't the best we can do, because it groups all stations together. Ideally we would create one model per station, with inputs from every station.

```{r lm.many, cache=TRUE, echo=FALSE}
# Construct a data frame with both long and wide properties. For every day for
# every station, we indicate spaces in that station at 0930, and spaces in all
# stations at 0900.
get.station <- function(short, long) {
  tmp <- spaces.0900.0930[spaces.0900.0930$name==long,
                          c('updated.date', 'at.0900')]
  names(tmp)[2] <- short
  return(tmp)
}
spaces.tmp <- merge(spaces.0900.0930, get.station('sp', 'Southampton Place'))
spaces.tmp <- merge(spaces.tmp, get.station('hh', 'High Holborn'))
spaces.tmp <- merge(spaces.tmp, get.station('bp', 'Bury Place'))
spaces.tmp <- merge(spaces.tmp, get.station('es', 'Earnshaw Street'))

summary(lm(at.0930 ~ sp+hh+bp+es,
           spaces.tmp[spaces.tmp$name=='Southampton Place',]))
summary(lm(at.0930 ~ sp+hh+bp+es,
           spaces.tmp[spaces.tmp$name=='High Holborn',]))
summary(lm(at.0930 ~ sp+hh+bp+es,
           spaces.tmp[spaces.tmp$name=='Bury Place',]))
summary(lm(at.0930 ~ sp+hh+bp+es,
           spaces.tmp[spaces.tmp$name=='Earnshaw Street',]))
```

Southampton Place has slightly regressed, but the others have improved slightly. In particular, Bury Place gets an R^2 of 0.89, which is pretty good. (It's important to note that this doesn't make our model worse for Southampton Place than the aggregate model. The aggregate model was just overconfident on that station.)

Final plots and summary
-----------------------

### Plot 1

```{r final.plot.1, echo=FALSE, warning=FALSE, cache=TRUE, fig.width=11, fig.height=11}
ggplot(mutate(bikes.wide, timestr=strftime(time, '%H:%M', tz='GMT')),
       aes(x=total)) + 
  geom_histogram(binwidth=1) + 
  facet_wrap(~timestr) +
  xlab('Total number of bikes available') +
  ylab('Frequency') +
  ggtitle('Number of bikes available throughout the day')
```

The total number of bikes available changes gradually throughout the day, with few bikes typically available at night, but often many available during the daytime. The distribution looks left-skewed from around 10:00 to 16:30, and right-skewed from around 19:00 to 07:30. The left skew is never as extreme as the right skew, but because the stations have different numbers of slots, that doesn't tell us much.

### Plot 2

```{r final.plot.2, echo=FALSE, warning=FALSE, cache=TRUE, fig.width=11, fig.height=11}
ggplot(subset(spaces.0900.0930, is.weekday),
       aes(x=at.0900, y=at.0930, color=name)) +
  geom_jitter(alpha=0.3, position=position_jitter(width=0.2, height=0.2)) +
  stat_smooth(alpha=0.1, linetype='dashed') +
  scale_x_continuous(breaks=seq(0, 22)) +
  scale_y_continuous(breaks=seq(0, 22)) +
  xlab('Spaces available at 09:00') +
  ylab('Spaces available at 09:30') +
  ggtitle('Spaces available during the week at 09:00 and 09:30, by station') +
  geom_line(data=data.frame(x=c(0,21), y=c(0, 21)),
            aes(x=x, y=y),
            color='#555555', linetype='dashed') +
  guides(color=guide_legend(title='Station'))
```

This time around, I restricted the graph to weekdays only. It's rare for the number of stations to go up between 09:00 and 09:30. All four stations have similar usage patterns.

At 09:00, if there are five or fewer spaces available, it looks as though the most common single outcome at 09:30 is no spaces at all.

Points above the dotted black line are ones where more spaces were available at 09:30 than at 09:00. (Caveat: I've applied slight jittering, so points very close to that line are ones where the same number of spaces were available.) There are obviously much fewer of them. However, the top-left corner of the graph has a few points in it where the bottom-right corner is empty. The number of bikes never goes down by more than eleven, but it goes up by as much as fifteen.

### Plot 3

```{r final.plot.3, echo=FALSE, warning=FALSE, cache=TRUE, fig.width=11, fig.height=11}
# Get a data frame to easily plot the number of docking points at each station.
docks <- merge(group_by(bikes, name) %>% summarise(docks=max(num.docks)),
               data.frame(x=as.time(c(0, 86399))))

ggplot(bikes, aes(x=updated.time.bin, y=num.bikes, color=name)) +
  geom_line(aes(linetype='1'), stat='summary', fun.y=mean) +
  geom_line(aes(linetype='2'), stat='summary', fun.y=median, alpha=0.5) +
  geom_line(aes(linetype='3'), stat='summary', fun.y=quantile, probs=0.9) +
  geom_line(aes(linetype='3'), stat='summary', fun.y=quantile, probs=0.1) +
  geom_line(aes(x=x, y=docks, color=name, linetype='4'),
            alpha=0.5, data=docks) +
  scale_x_datetime(breaks=as.POSIXct(seq(0, 86400, 3600*3),
                                     origin='1970-01-01'),
                   labels=date_format('%H:%M')) +
  xlab('Time') + ylab('Bikes available') +
  ggtitle('Number of bikes available by time and station') +
  scale_linetype_manual(name='Statistic',
                        values=c('solid', 'dashed', 'dotted', '15'),
                        labels=c('Mean', 'Median', '10th/90th percentile',
                                 'Total number of docks')) +
  guides(color=guide_legend(title='Station'))
```

I took advantage of binning to calculate specific summary functions. All stations show similar patterns: at night, there are few bikes available; during office hours, there are almost always some, and the 10-90 percentile range is a lot higher. The trough around 1pm in the previous version of this plot no longer shows up, which makes me suspect it was simply an artifact of the smoothing method.

During the day, the number of bikes available is generally ranked by the number of docking slots at each station - so High Holborn has the least, and Bury Place has the most. When the bikes are taken around 18:00, High Holborn seems to lose them more slowly than the other stations. For Earnshaw Street and especially Bury Place, the 90th percentile lines suggest that those two stations were often completely full.

Reflection
----------

I've learned a lot about how to fight ggplot when it doesn't do exactly what I want by default, and in particular about how to shape my data for it.

I feel like a data frame isn't an ideal structure for the data I have. The fact that I had to create `prev.*` and `d.*` copies of those columns that need it seems suboptimal, ideally I would have wanted to be able to refer directly to offset rows in the data. (For example, there's currently no easy way to ask "what's the difference between the number of bikes now and 30 minutes ago?") But I couldn't find anything that worked better. In particular, time series only allow one data type, so I would have had to fight to use them at all, and I don't know if they would have been any more useful.

My data set itself isn't ideal, particularly in the amount of missing data. Unfortunately, I don't think any better historical bike record data is available. I think I have enough data to trust my conclusions.

In general, it seems that weather doesn't have much impact on bike usage. I checked rain, fog, temperature and wind speed, and only rain made a significant difference. But since the rainfall data seems to be internally inconsistent, I don't know how much we can learn from it. It would be useful to validate it from another source. We might also learn more with finer-grained weather data. For example, when predicting bike availability at a specific time, it doesn't help much if we know whether or not it rained at all on a given day; but it might help more to know whether it was raining at that particular time.

On the other hand, we can make pretty good predictions about future bike (and slot) availability just from current availability. An ambitious future project might be a prediction system. A user could specify a station and an arrival time, and the system could tell her how likely it would be that she could find a slot in that station and nearby ones, and suggest an earlier arrival time that would increase that chance.

One thing I didn't examine was [public holidays](http://en.wikipedia.org/wiki/Bank_holiday#List_of_current_holidays_in_the_United_Kingdom.2C_Ireland_and_the_Isle_of_Man). For example, we might ask whether, on plot 2 above, many of the points where spaces were freed up fell on holidays. (We can calculate 85 points above the line, and only 8*4 = 32 of them could be on public holidays, but that's still potentially a third of them.)

After initially submitting this report, I noticed a big problem. All timestamps were collected and reported in physical time, but bike usage patterns are going to be related to clock time. So some of my graphs, particularly later ones, were mixing in data from two different clock times (e.g. 09:00 and 10:00) as if they were the same. I've corrected that error in this version.
